---
title: 'Dataset Design: Temporal Concurrency - What'
author: "Chionesu George"
date: 2023-12-26
date-modified: last-modified
slug: what
series: 
  - Dataset Design and Temporal Concurrency
categories:
  - Data Engineering
  - Theory
tags:
  - analytics
  - engineering
  - time
  - data
  - semantics
editor_options: 
  chunk_output_type: console
---

{{% blogdown/jquery %}}

```{r SETUP, echo=FALSE}
#| label: SETUP
#| echo: false

knitr::opts_chunk$set(echo = FALSE);

options(article="dataset-design-and-temporal-concurrency");

source(file = "setup.R", local = environment());

tags$link(rel="stylesheet", href="/markdown.css");

tags$script(src = "/markdown.js");

# last_mod

tags$span(style = "display: hidden;", taxonomy);

hourglass.png <- magick::image_read("C:/Users/sapie/GitHub/delriaan.github.io/static/hourglass.png");

break_time <- tags$p(
  tags$hr(style = "border-bottom: solid 5px #555555; width:100%; ")
  , tags$span(
      style = "font-size:4em;"
      , class = "speech"
      , tags$img(
          src = "/hourglass.png"
          , caption="Take a Break!"
          , style="height:3em; vertical-align:text-top; margin-top:-50px; float:left"
          )
      , "Take a Break!"
      )
  , tags$hr(style = "border-top: solid 5px #555555; width:100%; ")
  )

```

[Welcome back!]{.decorativeText} 

In [Part 1](../who-and-when){target="blank"}, we discussed the importance of giving proper treatment to understanding the *who* and *when* of a problem statement which provide the context within which a solution will be crafted. In this article, we'll discuss the quantitative building blocks of the problem statement itself.

## Metrics & Dimensions

```{r}
drop_hint("First, let's define the key measures and related metrics listed in the problem statement:", context = "posthoc", toggleGroup = 1) 
```

- [ $\gamma_1$]{.bigMath}: [Average length of stay]{#msg_1} $\\\gamma(x)=\bar{x};\enspace x := \big(t_{i+1}-t_{i}\big)$

- [ $\gamma_2$]{.bigMath}: [Counts of lapses in medication adherence]{#msg_2} $\\f(x)=\sum{x};\enspace x:=\Big\{\matrix{0, \text{No lapse}\\1, \text{Lapse}}$ (there's more to this operation which will be covered in Part 3)
  
- [ $\gamma_3$]{.bigMath}: [Cumulative count of lapses in medication adherence]{#msg_3} $\\f(\gamma_2, j)=\sum_{i=1}^{j}{\gamma_2|j\le{i}}$ (yes, I have not defined $j$: that will be covered in Part 3)

- [ $\gamma_{4}$]{.bigMath}: [Number of unique members]{#msg_4} $\\f(x)=\#x;\enspace x:=\text{the set of unduplicated member identifiers}$
  
- [ $\gamma_5$]{.bigMath}: [Total expenditures]{#msg_5} $\\f(x)=\sum{x};\enspace x:=\text{cost}$

### Metrics vs. Measures

Note that for each of metrics, a *metric* and a *measure* were defined. The measure is the content of the metric, while the metric operates on the measure.

>
For example, [ $\gamma_1$]{msg_id=1}'s measure is *days between dates* and the metric is *mean of {measure}*.

### Dimensions

The dimensions of the problem statement are the *who* and *when* of the problem statement. These were discussed in [Part 1](../who-and-when){target="_blank"}.


{{% blogdown/footer %}}